---
title: "Capstone Tables 1, 2 & 3"
author: "Gift Kiti"
date: "2/10/2020"
output: html_document
---

# Creating Descriptive Tables and Exploring Data (continued)

Here I will be creating descriptive analyses and cross-tabulatons to build Tables 1 and 2 for my thesis research project looking at the association between having a companion during labor and person centered maternity care. In addition, I will run some statistical tests to further explore relationships between categorical/continuous variables. 

# Setting my work directory:
1) Go to Session -> Set Working Directory -> Choose Directory...OR 2) Locate your file path, copy and paste it within the function setwd("filepath")
```{r}
setwd("~/Dropbox/Gift captstone/Data")
```

## Call in the dataset I created last week
```{r}
#read in the dataset created from the previous week 
pqcc_raw_3 <- read.csv("final_data.csv")
#pqcc_raw_3 <- data.frame(pqcc_raw_3)

```

```{r r loadlib, echo=T, results='hide', message=F, warning=F}
library(summarytools) #descriptve statistics / cross-tabulations

#install.packages("ggplot2")
library(ggplot2)
```

# Summary Tables - Descriptive Analysis (TABLE 1)

In this section, I am going to generate descriptive statistics using both the 'freq' function and the 'c' function.  The package, summarytools, is fairly versatile. You can see more examples and techniques here: https://cran.r-project.org/web/packages/summarytools/vignettes/Introduction.html


## Method 1 - 'freq' function, repeated
```{r}
freq(pqcc_raw_3$marital_stat3, cumul = FALSE)
freq(pqcc_raw_3$work_for_pay, cumul = FALSE)
freq(pqcc_raw_3$num_births_cat2, cumul = FALSE)
freq(pqcc_raw_3$NationalQuintile_3cat, cumul = FALSE)
#freq(pqcc_raw_3$del_place, cumul = FALSE)
#freq(pqcc_raw_3$del_place1, cumul = FALSE)
freq(pqcc_raw_3$del_place2, cumul = FALSE)
freq(pqcc_raw_3$company_tohf, cumul = FALSE) 
#freq(pqcc_raw_3$education2, cumul = FALSE) 
freq(pqcc_raw_3$education, cumul = FALSE) 
#freq(pqcc_raw_3$grade_completed, cumul = FALSE) 
#freq(pqcc_raw_3$del_provider1, cumul = FALSE) 
freq(pqcc_raw_3$del_provider4, cumul = FALSE)
freq(pqcc_raw_3$del_provider_sex, cumul = FALSE) 
#freq(pqcc_raw_3$tribe, cumul = FALSE) 
freq(pqcc_raw_3$tribe1, cumul = FALSE)  
#freq(pqcc_raw_3$religion, cumul = FALSE) 
#freq(pqcc_raw_3$religion1, cumul = FALSE) 
freq(pqcc_raw_3$religion3, cumul = FALSE) 
freq(pqcc_raw_3$interview_location, cumul = FALSE)
#freq(pqcc_raw_3$postpartum_length_bin, cumul = FALSE)
freq(pqcc_raw_3$postpartum_length_bin2, cumul = FALSE)
#freq(pqcc_raw_3$pcmc_short_score, cumul = FALSE)
freq(pqcc_raw_3$pcc_support_binary, cumul = FALSE)



#Who accompanied to health facility
#Here I'm treating each of the people that accompanied as a separate variable
freq(pqcc_raw_3$comp_tohf_type___1, cumul= FALSE)
freq(pqcc_raw_3$comp_tohf_type___2, cumul= FALSE)
freq(pqcc_raw_3$comp_tohf_type___3, cumul= FALSE)
freq(pqcc_raw_3$comp_tohf_type___4, cumul= FALSE)
freq(pqcc_raw_3$comp_tohf_type___5, cumul= FALSE)
#freq(pqcc_raw_3$comp_tohftype_other, cumul= FALSE) cowife seems to have accompanied a good number of the people

#there are a variety of settings for this function, freq. 
#type in "?freq" to see different settings. For example, report.NAs = FALSE
```
```{r}
#Short PCMC scale
#pcmc_short <- pqcc_raw %>% select(pcc_name, pcc_respect, pcc_friendly,
#pcc_explain_exams, pcc_explain_meds, pcc_questions, pcc_involvement, pcc_permission, pcc_position_choice, pcc_attention_help, pcc_feeling,  pcc_privacy_vis_rc,  pcc_bestcare)
freq(pqcc_raw_3$pcc_friendly, cumul= FALSE)
freq(pqcc_raw_3$pcc_respect, cumul= FALSE)
freq(pqcc_raw_3$pcc_explain_exams, cumul= FALSE)
freq(pqcc_raw_3$pcc_explain_meds, cumul= FALSE)
freq(pqcc_raw_3$pcc_questions, cumul= FALSE)
freq(pqcc_raw_3$pcc_position_choice, cumul= FALSE)
freq(pqcc_raw_3$pcc_involvement, cumul= FALSE)
freq(pqcc_raw_3$pcc_permission, cumul= FALSE)
freq(pqcc_raw_3$pcc_attention_help, cumul= FALSE)
freq(pqcc_raw_3$pcc_feeling, cumul= FALSE)
freq(pqcc_raw_3$pcc_bestcare, cumul= FALSE)
freq(pqcc_raw_3$pcc_privacy_vis_rc, cumul= FALSE)
freq(pqcc_raw_3$pcc_name, cumul= FALSE)
freq(pqcc_raw_3$company_tohf, cumul= FALSE)


#Who accompanied to health facility
#Here I'm treatating each of the people that accompanied as a separate variable
freq(pqcc_raw_3$comp_tohf_type___1, cumul= FALSE)
freq(pqcc_raw_3$comp_tohf_type___2, cumul= FALSE)
freq(pqcc_raw_3$comp_tohf_type___3, cumul= FALSE)
freq(pqcc_raw_3$comp_tohf_type___4, cumul= FALSE)
freq(pqcc_raw_3$comp_tohf_type___5, cumul= FALSE)
#freq(pqcc_raw_3$comp_tohftype_other, cumul= FALSE) cowife seems to have accompanied a good number of the people

```

## Method 2 - Using 'c' function

This function "c" is a generic function which combines arguments in to a list. It is simple function, but extremely useful.

```{r results='hide'}
#Relevant Variables: subsetting to variables I am interested in (check spelling!)
#note: R is a Case Sensitive Language
t1_vars <-pqcc_raw_3[c("education", "marital_stat3", "occupation","age_cat", "work_for_pay", "NationalQuintile_3cat", "num_births_cat2", "education2", "del_provider_sex","num_births_cat2", "del_provider4", "del_place2", "tribe1", "religion3","interview_location","postpartum_length_bin2", "pcc_support_binary","company_tohf")]
t1_vars
#comp_tohf_type"- who accompanied you


```


##Example of Table 2: Descriptive Statistics by Exposure (labor companion)

## Categorical - Using the built=in chisq.test function
```{r results='hide'}
# Method 1 - No Ns, just chi-square tests
chisq.test(pqcc_raw_3$age_raw, pqcc_raw_3$pcc_support_binary, correct=FALSE) #correct=FALSE- continuity correction
chisq.test(pqcc_raw_3$education, pqcc_raw_3$pcc_support_binary, correct=FALSE)
chisq.test(pqcc_raw_3$marital_stat3, pqcc_raw_3$pcc_support_binary, correct=FALSE)
chisq.test(pqcc_raw_3$del_place2, pqcc_raw_3$pcc_support_binary, correct=FALSE)

#same line as earlier
t1_vars <-pqcc_raw_3[c("age_raw", "education", "marital_stat3", "del_place2")]

#applying this list to generate chi-squares at once
lapply(t1_vars, function(x) chisq.test(table(x,pqcc_raw_3$pcc_support_binary), #lapply-list apply
                            simulate.p.value = TRUE, correct=FALSE))
```


## Cross-tabulation and Chi-Squares using 'ctable'

'ctable' is a function in summarytools that produces a cross-tabulation for a pair of categorical variables (or factors) with either row, column, or total proportions, as well as marginal sums. Be careful with calculating chi-squares for variables with missing rows.

```{r results='hide'}
# Method 2 with Ns -- COMPARE CHI SQUARES

#c("literacy_read", "literacy_write", "grade_completed", "marital_status", "occupation", "part_education", "work_for_pay", "NationalQuintile_3cat", "num_births_cat2", "education2", "del_provider_sex", "del_provider4", "del_place2", "tribe1", "religion3","interview_location","postpartum_length_bin" )]

ctable(pqcc_raw_3$age_raw, pqcc_raw_3$pcc_support_binary, chisq = T, useNA = "no") 

#see how the chisquare changes when you don't include useNA
ctable(pqcc_raw_3$age_raw, pqcc_raw_3$pcc_support_binary, chisq = T) #didn't see a difference in the 2 tables


ctable(pqcc_raw_3$marital_stat3, pqcc_raw_3$pcc_support_binary, chisq = T, useNA = "no")
ctable(pqcc_raw_3$pcc_support_binary, pqcc_raw_3$pcc_support_binary, chisq = T, useNA = "no")
#ctable(pqcc_raw_3$literacy_read, pqcc_raw_3$pcc_support_binary, chisq = T, useNA = "no")
ctable(pqcc_raw_3$education, pqcc_raw_3$pcc_support_binary, chisq = T, useNA = "no")
ctable(pqcc_raw_3$work_for_pay, pqcc_raw_3$pcc_support_binary, chisq = T, useNA = "no")
ctable(pqcc_raw_3$NationalQuintile_3cat, pqcc_raw_3$pcc_support_binary, chisq = T, useNA = "no")
ctable(pqcc_raw_3$num_births_cat2, pqcc_raw_3$pcc_support_binary, chisq = T, useNA = "no")
ctable(pqcc_raw_3$interview_location, pqcc_raw_3$pcc_support_binary, chisq = T, useNA = "no")#0 is health facility and 1 is community or home
ctable(pqcc_raw_3$postpartum_length_bin2, pqcc_raw_3$pcc_support_binary, chisq = T, useNA = "no")
ctable(pqcc_raw_3$tribe1, pqcc_raw_3$pcc_support_binary, chisq = T, useNA = "no")
ctable(pqcc_raw_3$del_provider_sex, pqcc_raw_3$pcc_support_binary, chisq = T, useNA = "no")
ctable(pqcc_raw_3$del_provider4, pqcc_raw_3$pcc_support_binary, chisq = T, useNA = "no")
ctable(pqcc_raw_3$del_place2, pqcc_raw_3$pcc_support_binary, chisq = T, useNA = "no")
ctable(pqcc_raw_3$company_tohf, pqcc_raw_3$pcc_support_binary, chisq = T, useNA = "no")



#company to health facility
chisq.test(pqcc_raw_3$comp_tohf_type___1, pqcc_raw_3$pcc_support_binary, correct=FALSE)
chisq.test(pqcc_raw_3$comp_tohf_type___2, pqcc_raw_3$pcc_support_binary, correct=FALSE)
chisq.test(pqcc_raw_3$comp_tohf_type___3, pqcc_raw_3$pcc_support_binary, correct=FALSE)
chisq.test(pqcc_raw_3$comp_tohf_type___4, pqcc_raw_3$pcc_support_binary, correct=FALSE)
chisq.test(pqcc_raw_3$comp_tohf_type___5, pqcc_raw_3$pcc_support_binary, correct=FALSE)

#library(forcats)
```

Using T-test: We use t-test when we are dealing with 1 categorical variable(2 levels) and 1 continuous variable
```{r results='hide'}
#t-TEST
t.test((pqcc_raw_3$age)~(pqcc_raw_3$pcc_support_binary)) # where y is numeric and x is a binary factor
t.test((pqcc_raw_3$num_births)~(pqcc_raw_3$pcc_support_binary))

#ctable(t1, df$ag_work, chisq = T) DOES NOT WORK

#mean Maternal Age by exposure
#pqcc_raw_3 %>%
#group_by(pcc_support_binary) %>%
#summarise(mean_age = mean( age ),
#sd_vol = sd(age ),
#samp_size = n())
```
## Continuous variables

For continous variables, you will almost always run a simple t. test first.

USAGE:
t.test(x, y = NULL,
       alternative = c("two.sided", "less", "greater"),
       mu = 0, paired = FALSE, var.equal = FALSE,
       conf.level = 0.95, ...)

##Table 3: Descriptive Statistics by Outcome (PCMC scale)
In computing the p value for table 3 we shall either carry out an anova test or a t test. We will carry out an anova test if we are dealing with 1 variable that is continuous and another one that is categorical in 2 levels. T-test is used when we are dealing with 1 continuous variable and a categorical variable that only has 2 levels.
### Continuous - One-way ANOVA- >2 levels (continuous)

Null hypothesis: the means of the different groups are the same
Alternative hypothesis: At least one sample mean is not equal to the others.

```{r}
# Compute the analysis of variance

lm.test <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$age_raw))
# Summary of the analysis
summary(lm.test)
anova(lm.test)

# calculate the means and SD for each group using dplyr, then anova test if categorical variable is > 2 levels

#del_provider_sex

lm.test1 <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$del_provider3))
# Summary of the analysis
summary(lm.test1)
anova(lm.test1)

#marital status
pqcc_raw_3 %>%
group_by( marital_stat3) %>%
summarise(mean_score = mean( pcmc_short_score ),
sd_vol = sd( pcmc_short_score ),
samp_size = n())

lm.test2 <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$marital_stat3))
# Summary of the analysis
summary(lm.test2)
anova(lm.test2)

#age_cat
pqcc_raw_3 %>%
group_by(age_cat) %>%
summarise(mean_age = mean(pcmc_short_score),
sd_vol = sd(pcmc_short_score),
samp_size = n())

lm.test3 <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$age_cat))
# Summary of the analysis
summary(lm.test3)
anova(lm.test3)

#literacy
pqcc_raw_3 %>%
group_by(literate) %>%
summarise(mean_score = mean(pcmc_short_score),
sd_vol = sd(pcmc_short_score),
samp_size = n())

lm.test4 <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$literate))
# Summary of the analysis
summary(lm.test4)
anova(lm.test4)

#maternal education

pqcc_raw_3 %>%
group_by(education) %>%
summarise(mean_score = mean(pcmc_short_score),
sd_vol = sd(pcmc_short_score),
samp_size = n())

lm.test5 <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$education))
# Summary of the analysis
summary(lm.test5)
anova(lm.test5)

#wealth quintile
pqcc_raw_3 %>%
group_by(NationalQuintile_3cat) %>%
summarise(mean_score = mean(pcmc_short_score),
sd_vol = sd(pcmc_short_score),
samp_size = n())

lm.test6 <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$NationalQuintile_3cat))
# Summary of the analysis
summary(lm.test6)
anova(lm.test6)

#Work for pay( 2-levels)

#t-TEST
t.test((pqcc_raw_3$pcmc_short_score)~(pqcc_raw_3$work_for_pay)) # where y is numeric and x is a binary factor

pqcc_raw_3 %>%
group_by(work_for_pay) %>%
summarise(mean_score = mean(pcmc_short_score),
sd_vol = sd(pcmc_short_score),
samp_size = n())

lm.test7 <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$work_for_pay))
# Summary of the analysis
summary(lm.test7)
anova(lm.test7)

#number of births num_births_cat2

pqcc_raw_3 %>%
group_by(num_births_cat2) %>%
summarise(mean_score = mean(pcmc_short_score),
sd_vol = sd(pcmc_short_score),
samp_size = n())

lm.test8 <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$num_births_cat2))
# Summary of the analysis
summary(lm.test8)
anova(lm.test8)

#interview location 

pqcc_raw_3 %>%
group_by(interview_location) %>%
summarise(mean_score = mean(pcmc_short_score),
sd_vol = sd(pcmc_short_score),
samp_size = n())

#t-TEST
t.test((pqcc_raw_3$pcmc_short_score)~(pqcc_raw_3$interview_location)) # where y is numeric and x is a binary factor

lm.test9 <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$interview_location))
# Summary of the analysis
summary(lm.test9)
anova(lm.test9)

#postpartum length 
pqcc_raw_3 %>%
group_by(postpartum_length_bin) %>%
summarise(mean_score = mean(pcmc_short_score),
sd_vol = sd(pcmc_short_score),
samp_size = n())
#t-TEST
t.test((pqcc_raw_3$pcmc_short_score)~(pqcc_raw_3$postpartum_length_bin2)) # where y is numeric and x is a binary factor

lm.test10 <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$postpartum_length_bin2))
# Summary of the analysis
summary(lm.test10)
anova(lm.test10)

#tribe 
pqcc_raw_3 %>%
group_by(tribe1) %>%
summarise(mean_score = mean(pcmc_short_score),
sd_vol = sd(pcmc_short_score),
samp_size = n())

lm.test11 <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$tribe1))
# Summary of the analysis
summary(lm.test11)
anova(lm.test11)

#delivery facility type del_place2
pqcc_raw_3 %>%
group_by(del_place2) %>%
summarise(mean_score = mean(pcmc_short_score),
sd_vol = sd(pcmc_short_score),
samp_size = n())

lm.test12 <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$del_place2))
# Summary of the analysis
summary(lm.test12)
anova(lm.test12)

#delivery provider4
pqcc_raw_3 %>%
group_by(del_provider4) %>% summarise(mean_score = mean(pcmc_short_score),
sd_vol = sd(pcmc_short_score),
samp_size = n())

lm.test13 <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$del_provider4))
# Summary of the analysis
summary(lm.test13)
anova(lm.test13)

mean(pqcc_raw_3$pcmc_short_score)
sd(pqcc_raw_3$pcmc_short_score)

#Comp_to_hf
pqcc_raw_3 %>%
group_by(company_tohf) %>%
summarise(mean_score = mean(pcmc_short_score),
sd_vol = sd(pcmc_short_score),
samp_size = n())
#t-TEST
t.test((pqcc_raw_3$pcmc_short_score)~(pqcc_raw_3$company_tohf)) # where y is numeric and x is a binary factor

#continuous support
library(dplyr)
pqcc_raw_3 %>%
group_by(pcc_support_binary) %>%
summarise(mean_score = mean(pcmc_short_score),
sd_vol = sd(pcmc_short_score),
samp_size = n())

#t-TEST
t.test((pqcc_raw_3$pcmc_short_score)~(pqcc_raw_3$pcc_support_binary)) # where y is numeric and x is a binary factor


lm.test14 <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$company_tohf))
# Summary of the analysis
summary(lm.test14)
anova(lm.test14)
#pqcc_raw_3$comp_tohf_type___1

#continuous labor support
pqcc_raw_3 %>%
group_by(pcc_support_binary) %>%
summarise(mean_score = mean(pcmc_short_score),
sd_vol = sd(pcmc_short_score),
samp_size = n())
#t-TEST
t.test((pqcc_raw_3$pcmc_short_score)~(pqcc_raw_3$pcc_support_binary)) # where y is numeric and x is a binary factor


lm.test15 <- lm(pqcc_raw_3$pcmc_short_score~ as.factor(pqcc_raw_3$pcc_support_binary))
# Summary of the analysis
summary(lm.test15)
anova(lm.test15)
```

##For Age and parity as continuous variables with outcome

Pearson’s correlation - measures a linear dependence between two variables cor(x, y, method = c(“pearson”, “kendall”, “spearman”)) cor.test(x, y, method=c(“pearson”, “kendall”,“spearman”))cor() computes the correlation coefficient cor.test() test for association/correlation between paired samples. It returns both the correlation coefficient and the significance level(or p-value) of the correlation.

```{r}
cor.test(pqcc_raw_3$age, pqcc_raw_3$pcmc_short_score, method=c("pearson"))
cor.test(pqcc_raw_3$num_births, pqcc_raw_3$pcmc_short_score, method=c("pearson"))
```

```{r}
# Save our final data-set as final data to be used in running multivariable models 
saveRDS(pqcc_raw_3, "pqcc_raw_3.rds")
# Restore it under a different name
final_data <- readRDS("pqcc_raw_3.rds")
```






