---
title: "Thesis data"
author: "Gift Kiti"
date: "2/19/2019"
output: pdf_document
---

The dataset we will be using in this lesson is a STATA file called 'pqcc_mothers_surveyv3-8-19-18.dta'. I've saved it in the Dropbox- Gift Capstone- Data folder. 

# Setting my work directory (3 options):

1) Go to Session -> Set Working Directory -> Choose Directory...

OR 

2) Locate your file path, copy and paste it within the function setwd("filepath")

OR 

3) File -> Import Dataset 
```{r}
#THIS IS MY DIRECTORY
setwd("~/Dropbox/Gift captstone/Data")

```

# Installing and loading my libraries 
(aka packages to run specific functions)

Below are the packages and librarries I will need for this exercise. 
```{r r loadlib, echo=T, results='hide', message=F, warning=F}
####INSTALL AND THEN LOAD LIBRARIES

#install.packages("haven")
#install.packages("Hmisc")
#install.packages("sjlabelled")
#install.packages("summarytools")
#install.packages ("data.table")
#install.packages ("XQuartz")
#install.packages("ggplot2")

library(ggplot2)
library(haven) #IMPORTS SPSS, STATA, AND SAS FILES
library(Hmisc)#for the fuction describe
library(sjlabelled)#ADDS LABELS TO VARIABLES / READS IN LABELS
library(summarytools) #Descriptve statistics
library(dplyr)
```

# Reading in dataset 
(unless you used the Import Dataset button earlier)

```{r}
#rm(list=ls())
pqcc_raw <- read_dta("pqcc_mothers_surveyv3-8-19-18.dta")
#1052 obs, 1059 variables

```

#Exploring dataset
```{r results='hide'}
head(pqcc_raw ) #show the first few lines of the data
str(pqcc_raw ) #shows the structure of the data and value labels
#head(pqcc_raw$var) #displays a specific variable's value labels
head(pqcc_raw$comp_labor) #Example 

describe(pqcc_raw) #another function to show a concise statistical description

```

# Exploring specific variables  --  Part 1

```{r eval=FALSE}

###MATERNAL AGE (CATEGORICAL VARIABLE)

#LOOK AT DATA
hist(pqcc_raw$age)
summary(pqcc_raw$age)
describe(pqcc_raw$age)

table(pqcc_raw$age, exclude=NULL) #include NAs
freq(pqcc_raw$age, cumul = FALSE)

```

# Generating new variables - Part 1
```{r}
#GENERATE AGERAW 
pqcc_raw$age_raw<-ifelse(15<=pqcc_raw$age & pqcc_raw$age<=19, 1,
                  ifelse(20<=pqcc_raw$age & pqcc_raw$age<=29, 2,
                    ifelse(30<=pqcc_raw$age & pqcc_raw$age<=48, 3,NA)))
table(pqcc_raw$age_raw, exclude=NULL)

#LABEL AGERAW
pqcc_raw$age_raw <-factor(pqcc_raw$age_raw, levels=c(1,2,3), 
labels=c("15-19", "20-29", "30-48"), ordered=F)#changed to T for ordered

pqcc_raw$age_raw <- set_label(pqcc_raw$age_raw, label = "Maternal age categories")

#CHECK LABELS
freq(pqcc_raw$age_raw)
mean(pqcc_raw$age)# meanage
summary(pqcc_raw$num_births)# mean parity
```


# Generating new variables - Part 2
```{r}

#marital_status3- Here I am combining the partnered/cohabiting category to single category along with the widowed and divorced/separated then having the married category as a separate entity alone.

#freq(pqcc_raw_3$marital_status)
pqcc_raw$marital_stat3 <- ifelse(pqcc_raw$marital_status==1,1,
                                   ifelse(pqcc_raw$marital_status==2,1,
                                          ifelse(pqcc_raw$marital_status==3,3,
                                            ifelse(pqcc_raw$marital_status==4,1,
                                                ifelse(pqcc_raw$marital_status==5,1,NA)))))
table(pqcc_raw$marital_stat3, exclude=NULL)
#LABEL Marital status
pqcc_raw$marital_stat3 <-factor(pqcc_raw$marital_stat3, levels=c(1,3),  labels=c("Not married", "Married"), ordered=F)
#not married here is equal to Single/Widowed/Separated/Divorced/Cohabiting

pqcc_raw$marital_stat3 <- set_label(pqcc_raw$marital_stat3, label = "Marital status")
#Check labels
freq(pqcc_raw$marital_stat3)
```

# Generating new variables - Part 3

```{r}
#num_births
# Here I am adding those with 4 or more births into one level so that level 5(5 or more births) is part of level 4

freq(pqcc_raw$num_births, cumul = FALSE)
freq(pqcc_raw$num_births_cat, cumul = FALSE)

pqcc_raw$num_births_cat2 <- ifelse(pqcc_raw$num_births_cat==1,1,
                                     ifelse(pqcc_raw$num_births_cat==2,2,
                                            ifelse(pqcc_raw$num_births_cat==3,3,
                                                   ifelse(pqcc_raw$num_births_cat==4,4,
                                                          ifelse(pqcc_raw$num_births_cat==5,4,NA)))))
table(pqcc_raw$num_births_cat2, exclude=NULL)

#LABEL Number of Births
pqcc_raw$num_births_cat2 <-factor(pqcc_raw$num_births_cat2, levels=c(1,2,3,4),  labels=c("1", "2","3","4 or more"), ordered=F)

pqcc_raw$num_births_cat2 <- set_label(pqcc_raw$num_births_cat2, label = "Number of Births")
#Check labels
freq(pqcc_raw$num_births_cat2, cumul = FALSE)

```
# Generating new variables - Part 4
## Here I will generate a new variable pcc_explain meds_1 which will have 3 factors instead of 4 factors. All the 4th option factors which translate to (Did not get any medicine) were moved into the 3rd category of yes all the time due to conservative coding.
```{r}
#Method 1
#recoding the 4 items of pcc_explain_meds- goes to 3 as done in validation paper
pqcc_raw$pcc_explain_meds_1 <- pqcc_raw$pcc_explain_meds
pqcc_raw$pcc_explain_meds_1[pqcc_raw$pcc_explain_meds_1==4] <-3

#Method 2
#recoding the 4 items of pcc_explain_meds- goes to 3 as done in validation paper      
pqcc_raw$pcc_explain_meds_2 <- NA #creating an empty variable then populating it in the next step

pqcc_raw$pcc_explain_meds_2 <- ifelse(pqcc_raw$pcc_explain_meds == 0, 0,
                                  ifelse(pqcc_raw$pcc_explain_meds==1, 1,
                                      ifelse(pqcc_raw$pcc_explain_meds==2, 2, 
                                          ifelse(pqcc_raw$pcc_explain_meds ==3, 3, 
                                              ifelse(pqcc_raw$pcc_explain_meds== 4,3,NA)))))
table(pqcc_raw$pcc_explain_meds_2, exclude=NULL)

pqcc_raw$pcc_explain_meds_1 <- pqcc_raw$pcc_explain_meds_2

#how to delete a column
pqcc_raw$pcc_explain_meds_2 <- NULL
```


# Generating outcome variable 
```{r eval=FALSE}
###OUTCOME VARIABLE - PCMC Score

#Creating the PCMC_short scale using relevant variables and calling the new variable pcmc_short_score
pqcc_raw <- pqcc_raw %>% rowwise() %>% mutate(pcmc_short_score = sum(pcc_name, pcc_respect, pcc_friendly,
pcc_explain_exams, pcc_explain_meds_1, pcc_questions, pcc_involvement, pcc_permission, pcc_position_choice, pcc_attention_help, pcc_feeling,  pcc_privacy_vis_rc,  pcc_bestcare, na.rm = F))
table(pqcc_raw$pcmc_short_score , exclude=NULL) #include NAs
table(pqcc_raw$pcc_explain_meds, exclude=NULL)
#pcc_explain_meds- goes to number 4
#is.na(pqcc_raw$pcmc_short_score)
#pqcc_raw$pcc_name

summary(pqcc_raw$pcmc_short_score)
```

# Exploring specific variables - Part 1
```{r eval=FALSE}
###OUTCOME VARIABLE - PCMC Score
#LOOK AT DATA FOR THE VARIABLE YOU ARE RECODING

#Histogram
hist(pqcc_raw$pcmc_short_score)

#Shows the mean,median,25th and 75th quartiles,min,max
summary(pqcc_raw$pcmc_short_score, exclude=NULL) 

#more descriptive information 
describe(pqcc_raw$pcmc_short_score)

#simple frequency table (no percentages)
table(pqcc_raw$pcmc_short_score, exclude=NULL) # why is it including null?

freq(pqcc_raw$pcmc_short_score, cumul = FALSE) #what do % valid mean
#report.nas = FALSE will hide NAs and % Valid column
#remove cumul = FALSE if you want cumulative percentages

#set print function for copying and pasting purposes
print(freq(pqcc_raw$pcmc_short_score, cumul = FALSE), method = "render")

```
# Exploring specific variables  --  Part 2

```{r}
#Barplots
ggplot(pqcc_raw,aes(x = NationalQuintile_3)+
  geom_bar(aes(fill = type))+
  labs(y = "Frequency", x = "") +
theme_minimal(base_size = 15)+
  theme(legend.position = "top"))

table(pqcc_raw$age, exclude=NULL) #include NAs
freq(pqcc_raw$age, cumul = FALSE)

#To look at this visually
#Boxplots by 3 age categories
ggplot(data = pqcc_raw, aes(x = agecat, y = pcmc_short_score)) +
  geom_boxplot() 
#Boxplot with pcmc score on y axis
ggplot(data = pqcc_raw, aes( y = pcmc_short_score)) +
  geom_boxplot() 
summary(pqcc_raw_3$pcmc_short_score)
#Boxplot separating those who had a companion and those that didn't
ggplot(data = pqcc_raw_3, aes(x = "", y = age)) +
         geom_boxplot() + facet_grid(. ~ as.factor(pcc_support_binary)) +
         labs(x = "")
#Vertical boxplot for pcmc score with extra labeling and fancy dots
boxplot(pqcc_raw$pcmc_short_score, axes = FALSE, staplewex = 1)
text(y = boxplot.stats(pqcc_raw$pcmc_short_score)$stats, labels = boxplot.stats(pqcc_raw$pcmc_short_score)$stats, x = 1.25)

# Piechart for labor support
pie(pqcc_raw_3$pcc_support_binary)
library(ggplot2)
# Example on how to create a piechart
simuldata <- data.frame(Category = LETTERS[1:3], 
                        value = c(196, 149,127))

piechart <- ggplot(simuldata, aes(x="", y=value, fill=Category)) +
  geom_bar(width=1, stat="identity") +
  coord_polar("y", start=0) +
  xlab("") +
  ylab("Value") +
  theme(axis.text.x=element_blank(),axis.ticks.x=element_blank())
piechart

```

#Generating EXPOSURE variable

**Here we are creating a binary exposure variable using a previous variable pcc_support_labor. Our binary level will combine levels 0 (No never) and 1( Yes, a few times) as the level 0=No then the 3 other levels("yes most of the time" (2), "yes all of the time" (3) and "I did not want someone to stay with me"(4) as our level 1 =Yes). Those that responded that I did not want someone to stay with me are part of the 3rd level(Yes all of the time). This is once again because of conservative coding and here we are making the assumption that had the person wanted someone to accompany them they they would have been allowed to have someone with them.**
```{r}
#EXPOSURE Variable:

#LABEL Support during labor
#pqcc_raw$pcc_support_labor_rc <-factor(pqcc_raw$pcc_support_labor, levels=c(0,1,2,3,4),labels=c("No, Never", "Yes, a few times","Yes, most of the time","Yes, all of the time", "I did not want someone to stay with me"), ordered=F)
#pqcc_raw$support_labor <- set_label(pqcc_raw$pcc_support_labor, label = "Companion allowed during labor")
#Check labels
#freq(pqcc_raw$pcc_support_labor, cumul = FALSE)

#New Exposure Variable Generated:pcc_support_binary
pqcc_raw$pcc_support_binary<- ifelse(pqcc_raw$pcc_support_labor== 0,0,
                                      ifelse(pqcc_raw$pcc_support_labor==1, 0,
                                             ifelse(pqcc_raw$pcc_support_labor==2, 2, 
                                                ifelse(pqcc_raw$pcc_support_labor==3, 2,
                                       ifelse(pqcc_raw$pcc_support_labor==4,2,NA)))))#record to 3 as was done in the validation
#LABEL Support during labor
pqcc_raw$pcc_support_binary <-factor(pqcc_raw$pcc_support_binary, levels=c(0,2),  labels=c("No", "Yes"), ordered=F)

pqcc_raw$pcc_support_binary <- set_label(pqcc_raw$pcc_support_binary, label = "Companion allowed during labor")
#Check labels
table(pqcc_raw$pcc_support_binary, exclude=NULL)
#table(pqcc_raw$pcc_support_labor, exclude=NULL)-
freq(pqcc_raw$pcc_support_labor, cumul = FALSE)
freq(pqcc_raw$pcc_support_binary)
#describe(pqcc_raw$pcc_support_binary)

```


# Exclusion Criteria- Subsetting my data!

Here I will apply the exclusion criteria to determine the final sample size I will be working with. I will be creating a new data.frame for each time I do a subset so I can track my numbers
```{r results='hide'}
#Starting with total study population N= 1052
##KEEP ONLY WOMEN THAT DELIVERED IN HEALTH FACILITY
table(pqcc_raw$del_hf, exclude=NULL) #facility delivery are coded as 1
pqcc_raw_1 <-subset(pqcc_raw, del_hf==1)# remaining 894 (-158 non-health facility births)

###KEEP WOMEN WITH Complete data on PCMC score (Outcome)
summary(pqcc_raw$pcmc_short_score, exclude= NULL)
pqcc_raw_2<-subset(pqcc_raw_1, !is.na(pcmc_short_score))# remaining 881 (-13 missing outcome)

###KEEP WOMEN WITH Complete data on continuous labor support (Exposure)
table(pqcc_raw_2$pcc_support_labor, exclude=NULL) #N=1 NA
pqcc_raw_3 <-subset(pqcc_raw_2, !is.na(pcc_support_labor))
pqcc_raw_3 <-subset(pqcc_raw_3,(pcc_support_labor!=4))
# remaining 874 (-7 missing exposure)

###Final dataset: N=874 (dropped N=178)

#Lets save the data as final_data
write.csv(pqcc_raw_3,"final_datta.csv")


```









